---
title: 'Class 9: Breast cancer analysis with PCA and clustering'
author: "Karen Li"
date: "2/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Unsupervised Learning Mini-Project

**1. Exploratory Data Analysis**
```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data)
```

Let's see the first 6 rows of data
```{r}
head(wisc.df)
```

Number of patients is indicated by the number of rows
How many samples (ie patients) are in this dataset?

```{r}
nrow(wisc.df)
```

```{r}
wisc.data <- as.matrix(wisc.df[ ,3:32])
rownames(wisc.data) <- wisc.df$id
```

Finally, setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and  0 otherwise. Note that R coerces TRUE to 1 and FALSE to 0.

```{r}
wisc.df$diagnosis
```

How many cancer(m) and noncancer (b) samples do we have in our dataset?
```{r}
table(wisc.df$diagnosis)
```

```{r}
## diagnosis <- as.numeric(wisc.df$diagnosis)

## tmp <- rep(0, nrow(wisc.df))
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
```

Q1. How many observations are in this dataset?
```{r}
nrow(wisc.data)
```

Q2. How many variables/features in the data are suffixed with _mean?
```{r}
## colnames(wisc.data)

inds <- grep("_mean", colnames(wisc.data))

length(inds)
```

Q3. How many of the observations have a malignant diagnosis?
```{r}
sum(diagnosis)
```

**2. Principal Component Analysis**

```{r}
# check column means and standard deviation
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
#perform PCA on wisc.data by completing this code
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Let's see a plot of our results
```{r}
biplot(wisc.pr)
```

We need to make our own PCA plot

```{r}
plot(wisc.pr$x[ ,1], wisc.pr$x[ ,2], col=wisc.df$diagnosis, xlab ="PC1", ylab= "PC2")
```

Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
```{r}
plot(wisc.pr$x[ ,1], wisc.pr$x[ , 3], col=(diagnosis +1), xlab = "PC1", ylab = "PC3")
```

### Variance captured in each PC
```{r}
pr.var <- wisc.pr$sdev^2

# variance explained by each principal component
pve <- round((pr.var/sum(pr.var))*100, 1)
```

```{r}
# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2) )
```

**3. Hierarchical Clustering**

For hieracrchical clustering we need a few things:
1. Distance matrix **dist()** fucntion
2. the **hclust()** function
3. use the **cutree()** function to fund cluster membership vector

Scale wisc.data
```{r}
data.scaled <- scale(wisc.data)

round(apply(data.scaled, 2, sd), 1)
```

Calc the Euclidean distance between all pairs of observations in the scaled dataset and assign result to data.dist
```{r}
data.dist <- dist(data.scaled)
```

Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign result to wisc.hclust
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline (h=4, col="red", lty=2)
```

**5. Combining methods**

### Cluster in PCA space

For clustering we need:
1. distance matrix
2. clustering function
3. cutree

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis+1)
```

```{r}
pc.dist <- dist(wisc.pr$x[,1:2])
pc.hclust <- hclust(pc.dist, method="ward.D2")
plot(pc.hclust)
```

```{r}
grps3 <- cutree(pc.hclust, k=3)
table(grps3)
```

```{r}
table(grps3, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:3], col=grps3)
```


**7. Prediction**

We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis+1)
points(npc[,1], npc[,2], col="blue", pch=16, cex=2)
```

Q17. Which of these new patients should we prioritize for follow up based on your results?
The two blue points above represent our two new patients. The one with a larger PC1 value should be prioritized.


**Important functions used in this class**
grep ("_mean", x)
prcomp (x, scale = TRUE)
hclust (dist(x), method = "")














