---
title: 'Class 8: Clustering and PCA'
author: "Karen Li"
date: "2/5/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means clustering

```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```

```{r}
km <- kmeans(x, centers=2, nstart=20)
km
```
30 points in each cluster

tells us cluster size
```{r}
km$size
```

cluster assignment/mebership
```{r}
km$cluster
```

```{r}
km$center
```

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
```

```{r}
km$totss
```

## hierchical clustering

```{r}
# First we need to calculate point (dis)similarity
# as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
# clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc 
```

```{r}
plot(hc)
abline(h=6, col="red")
cutree(hc, h=6) # Cut by height h
```

We can also use k=number of cluster groups I want with **cutree()** 
```{r}
cutree(hc, k=2) #cut into k groups
```

### Different Linkage methods

```{r}
# using differenet hierchical clustering methods
d <- dist_matrix
hc.complete <- hclust(d, method="complete")
plot(hc.complete)

hc.average <- hclust(d, method="average")
plot(hc.average) 

hc.single <- hclust(d, method="single")
plot(hc.single)
```

Run hclust() on this made up overlapping data
```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```

Q. Use the dist(), hclust(), plot() and cutree()
 functions to return 2 and 3 clusters
Q. How does this compare to your known 'col' groups?

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
abline(h=2, col="red")
abline(h=2.5, col="blue")
```

```{r}
grp2 <- cutree(hc, k=2)
plot(x, col=grp2)
```


```{r}
grp3 <- cutree(hc, k=3)
plot(x, col=grp3)
```


## Principal Component Analysis 

```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1) 

head(mydata)
```

```{r}
## let's do PCA
pca <- prcomp(t(mydata), scale=TRUE)
pca
attributes(pca)
```

```{r}
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2]) 
```

```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2
## Precent variance is often more informative to look at
pca.var.per <- round(pca.var/sum(pca.var)*100, 1) 
pca.var.per
```

```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 

identify(pca$x[,1], pca$x[,2], labels=colnames(mydata)) 
```


# Worksheet
```{r}
x <- read.csv("UK_foods.csv")
```

```{r}
nrow(x)
ncol(x)
```


```{r}
x <- read.csv("UK_foods.csv", row.names=1)
head(x)
```

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

```{r}
pairs(x, col=rainbow(10), pch=16)
```

Let's run PCA on this to get some insight on the main differences between N. Ireland and te other UK countries

```{r}
pca <- prcomp(t(x))
summary(pca)
```

```{r}
# Plot PC1 vs PC2
plot(pca$x[ ,1], pca$x[ ,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x),
    col = c("orange", "red", "blue", "darkgreen"))
```

```{r}
v <- round(pca$sdev^2/sum(pca$sdev^2)*100)
v
```

```{r}
z <- summary(pca)
z$importance

barplot(v, xlab="Principal Component", ylab = "Percent Variation")
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

```{r}
## do same thing for PC2
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[ ,2], las=2 )
```

