---
title: "Class 6: Writing Your Own R Function"
author: "Karen Li"
date: "1/24/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Section 1: Reading Files

Here we are going to try to use the **read.table()** function and friends to read some example flat files.

First let's try adding a simple plot

```{r}
plot(1:10)
```

Back to file reading

```{r}
read.table("https://bioboot.github.io/bimm143_W19/class-material/test1.txt", header=TRUE, sep=",")
```

For this common CSV format, we can use **read.csv()** 

```{r}
file1 <- "https://bioboot.github.io/bimm143_W19/class-material/test1.txt"
data1 <- read.csv(file1)
data1
```

```{r}
file2 <- "https://bioboot.github.io/bimm143_W19/class-material/test2.txt"
read.csv(file2, sep="$")
```

```{r}
file3 <- "https://bioboot.github.io/bimm143_W19/class-material/test3.txt"
read.table(file3, sep="")
```

## Section 2: R Functions

My first function

```{r}
add <- function(x, y=1) {
  # Sum the input x and y
  x + y
}
```

Let's try using this function

```{r}
add(1)
```

```{r}
add(1,100)
```

```{r}
add(c(1,2,3))
```

```{r}
add( c(1, 2, 3), 4) 
```

```{r}
#add(1,2,3)
```

```{r}
#add(x=1, y="b")
```

My 2nd function

```{r}
rescale <- function(x) {
 rng <-range(x)
 (x - rng[1]) / (rng[2] - rng[1])
}
```

Testing on a small example where we know what the answer should be

```{r}
rescale(1:10)
```

```{r}
# How would you get your function to work here…
rescale( c(1,2,NA,3,10) )
```

```{r}

# rescale (c(1,10,"string”))
```

```{r}
rescale2 <- function(x) {
 rng <-range(x, na.rm=TRUE)
 (x - rng[1]) / (rng[2] - rng[1])
}

x <- c(1,2, NA,3,10)

rescale2(x)

```

```{r}
rescale3 <- function(x, na.rm=TRUE, plot=FALSE) {
  rng <-range(x, na.rm=na.rm)
  print("Hello")
  
  answer <- (x - rng[1]) / (rng[2] - rng[1])

  print("is it me you are looking for?")
 
 if(plot) {
 plot(answer, typ="b", lwd=4)
 }
 print("I can see it in ...")
 return(answer)
}
```

```{r}
rescale3(c(1:6, NA,10))
```


## Working with the bio3d package 

To install this package, I used the command **install.packages("bio3d")**.

```{r}
library(bio3d)
```

```{r}
# Read a PDB file from the database
s1 <- read.pdb("4AKE") # kinase with drug
s1
```

Trying out the provided code
```{r}
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

